---
  title: "gene expr - mut"
  output: html_notebook
---
  
```{r setup}
require(tidyverse)
require(fst)
require(furrr)
require(sparsebn)
require(tidyr)
require(dplyr)
# library(data.table)
library(dplyr)
library(stringr)
files <- list.files('../data/Lung', full.names = T, pattern = 'txt$') %>% 
  set_names(., str_replace(basename(.), '.txt',''))
files_fst <- map(files,str_c,'.fst')
```


```{r parameters}
# nr lines read from files (for testing)
n_lines <- 1000L

# size of 'active' interval [gene_start - M/2, gene_start + M/2] for each gene
M <- 10000

# number of most variable variables selected per layer
n_sel_var <- 1000

# sparsity
golden_lambda <- 5


```

# Data preparation

## Methylation data

```{r}
Lung_CNV <- read_fst(files_fst$Lung_CNV, to = n_lines) %>% as_data_frame
Lung_Methylation <- read_fst(files_fst$Lung_Methylation, to = n_lines) %>% as_data_frame %>% 
  # mutate(Gene = str_remove(Gene,'[,/!@#$%^&*()-]')) %>%  # remove invalid variable name characters
  distinct
  

# remove rows with NA (e.g. genes with no location)
Lung_CNV <- Lung_CNV %>% drop_na() 
```

Create Methylation data set. Aggregate by Gene/Chr (average) over 'active' interval gene start - M/2 to gene start + M/2

```{r}
# get probe_location
probe_location <-  
  Lung_Methylation %>% select(probeID, Chr, Gene, Start)  %>% 
  drop_na() %>% 
  mutate(Gene = sub(',.*','', Gene)) # create valid variable names

# join CNV gene location, and probe location
probe_gene_loc <- inner_join(
  Lung_CNV %>% select(Chr, Gene, Gene_Start = Start, Gene_Stop = Stop), 
  probe_location, 
  by = c("Chr", "Gene")) %>% 
  filter(Start > Gene_Start - M/2, Start < Gene_Stop + M/2)

# Select probes within gene interval

Lung_Methylation_Avg <- Lung_Methylation %>% 
  filter(probeID %in% probe_gene_loc$probeID) %>% 
  select(Chr, Gene, starts_with("TCGA")) %>% 
  drop_na(starts_with("TCGA")) %>% 
# get Average methylation level in each gene_chr
  group_by(Chr, Gene) %>% 
  summarize_at(.vars = vars(starts_with('TCGA')), .funs = mean, na.rm = TRUE) %>% 
  unite("gene_chr", Gene,Chr,sep = "_", remove = TRUE)
# 
Lung_Methylation_Avg_t <- Lung_Methylation_Avg %>% 
  gather(SampleID,methyl, starts_with('TCGA')) %>% 
  spread(gene_chr,methyl) %>% 
  rename_at(.vars = vars(-SampleID), .funs = str_c,"_methyl")

# files_fst$Lung_Methylation_t <- str_replace(basename(files_fst$Lung_Methylation), '([.].*)', '_t\\1')
# write_fst(Lung_Methylation_Avg_t, files_fst$Lung_Methylation_t)
``` 

## Gene expression data

```{r paged.print=TRUE}
Lung_GeneExpression_t <- read_fst(files_fst$Lung_GeneExpression, to = n_lines ) %>% as_data_frame %>% 
  gather(SampleID, value = gene_expression, starts_with("TCGA")) %>% 
  select(Gene,Chr,gene_expression,SampleID) %>%
  tidyr::unite(gene_chr, Gene,Chr) %>% 
  distinct(SampleID, gene_chr,.keep_all = TRUE) %>% 
  spread(gene_chr,gene_expression)  %>% 
  rename_at(.vars = vars(-SampleID), .funs = str_c,"_geneexpr")

#Lung_CNV <- read_fst("../../data/intermediates/Lung_CNV_t.fst") %>% as_data_frame
#Lung_miRNA <- read_fst("../../data/startset/Lung_miRNA.txt.fst") %>% as_data_frame
```

## Mutation data

Simply count the number of mutations per gene/chr 

```{r lung mutation data}
Lung_Mutation <- read_fst(files_fst$Lung_Mutation, to = n_lines) %>% as_data_frame
Lung_Mutation_t <-  
  Lung_Mutation %>%
  rename(SampleID = Sample_ID) %>% 
  drop_na(SampleID) %>% 
  count(SampleID, Gene, Chr) %>% 
  gather(SampleID, value = n, starts_with("TCGA")) %>%
  unite(gene_chr, Gene,Chr) %>%
  spread(gene_chr,n, fill = 0) %>%
  rename_at(.vars = vars(-SampleID), .funs = str_c,"_mut")
```

## CNV data

```{r}
Lung_CNV_t <-
  Lung_CNV %>% 
  gather(SampleID, value = cnv, starts_with("TCGA")) %>% 
  select(Gene,Chr,cnv,SampleID) %>%
  tidyr::unite(gene_chr, Gene,Chr) %>% 
  distinct(SampleID, gene_chr,.keep_all = TRUE) %>% 
  spread(gene_chr,cnv)  %>% 
  rename_at(.vars = vars(-SampleID), .funs = str_c,"_cnv")
```

# Dataset reduction heuristic

Create top1000 data frames

```{r select vars} 
var_mut <- apply(select(Lung_Mutation_t,-SampleID), 2, sd)
var_gen <- apply(select(Lung_GeneExpression_t, -SampleID), 2, sd)
var_meth <- apply(select(Lung_Methylation_Avg_t, -SampleID),2,sd)
var_cnv <- apply(select(Lung_CNV_t, -SampleID),2,sd)

n_sel_var_mut <- min(length(var_mut), n_sel_var)
n_sel_var_gen <- min(length(var_gen), n_sel_var)
n_sel_var_meth <- min(length(var_meth), n_sel_var)
n_sel_var_cnv <- min(length(var_cnv), n_sel_var)

vars_mut <- names(sort(var_mut, decreasing = TRUE)[1:n_sel_var_mut])
vars_gen <- names(sort(var_gen, decreasing = TRUE)[1:n_sel_var_gen])
vars_meth <- names(sort(var_meth, decreasing = TRUE)[1:n_sel_var_meth])
vars_cnv <- names(sort(var_cnv, decreasing = TRUE)[1:n_sel_var_cnv])


Lung_Mutation_t_sel <- Lung_Mutation_t %>% select(SampleID, vars_mut)
Lung_GeneExpression_t_sel <- Lung_GeneExpression_t %>% select(SampleID, vars_gen)
Lung_Methylation_t_sel <- Lung_Methylation_Avg_t %>% select(SampleID, vars_meth)
Lung_CNV_t_sel <- Lung_CNV_t %>% select(SampleID, vars_cnv)
```


# Covariance matrix

Sparse covariance between DNA layers (CNV, Mutation, Methylation) and RNA layer (Gene expression)


```{r settings for graph}
# create blacklist, NO LINKS inside same data set


``` 



# Generate Methyl -> RNA
```{r}
blacklist <- 
  rbind(
    expand.grid(1:n_sel_var_meth,1:n_sel_var_gen),
    expand.grid((n_sel_var_meth + 1):(n_sel_var_meth + n_sel_var_gen),(n_sel_var_meth + 1):(n_sel_var_meth + n_sel_var_gen))
  ) %>% as.matrix


meth_genexpr <- inner_join(Lung_Methylation_t_sel,Lung_GeneExpression_t_sel, by = "SampleID")

# golden_lambda <- 5
meth_genexpr_sparse <- sparsebnUtils::sparsebnData(meth_genexpr[,-1], type ="continuous")
cov_expr_methyl <- 
  sparsebn::estimate.covariance(
    meth_genexpr_sparse, 
    blacklist = blacklist, 
    lambdas = golden_lambda
  )

saveRDS(
  cov_expr_methyl,
  file= "../data/cov_expr_methyl.rds"
)
```

# Generate Mutation -> RNA
```{r}

mut_genexpr <-  inner_join(Lung_Mutation_t_sel,Lung_GeneExpression_t_sel, by = "SampleID")
blacklist <- 
  rbind(
    expand.grid(1:n_sel_var_mut,1:n_sel_var_gen),
    expand.grid((n_sel_var_mut + 1):(n_sel_var_mut + n_sel_var_gen),(n_sel_var_mut + 1):(n_sel_var_mut + n_sel_var_gen))
  ) %>% as.matrix

mut_genexpr_sparse <- sparsebnUtils::sparsebnData(mut_genexpr[,-1], type ="continuous")
cov_mut_genexpr <- 
  sparsebn::estimate.covariance(
    mut_genexpr_sparse, 
    blacklist = blacklist, 
    lambdas = golden_lambda
  )

saveRDS(
  cov_mut_genexpr,
  file="../data/cov_mut_genexpr.rds"
)

```


# Generate CNV -> RNA
```{r}
cnv_genexpr <-  inner_join(Lung_CNV_t_sel,Lung_GeneExpression_t_sel, by = "SampleID")
blacklist <- 
  rbind(
    expand.grid(1:n_sel_var_cnv,1:n_sel_var_gen),
    expand.grid((n_sel_var_cnv + 1):(n_sel_var_cnv + n_sel_var_gen),(n_sel_var_cnv + 1):(n_sel_var_cnv + n_sel_var_gen))
  ) %>% as.matrix
cnv_genexpr <- sparsebnUtils::sparsebnData(cnv_genexpr[,-1], type ="continuous")
cov_cnv_genexpr <- 
  sparsebn::estimate.covariance(
    cnv_genexpr, 
    blacklist=blacklist, 
    lambdas=golden_lambda
  )

saveRDS(
  cov_cnv_genexpr,
  file="../data/cov_cnv_genexpr.rds"
)
```

# Build full covariance matrix

```{r}
b <- cbind(matrix(0,ncol=1000,nrow=1000), cov_mut_genexpr[[1]][1:1000,1001:2000])
d <- cbind(rbind(cov_mut_genexpr[[1]][1:1000,1:1000],t(b)), rbind(b,cov_cnv_genexpr))

x <- rbind(matrix(0,ncol=1000,nrow=2000), cov_expr_methyl[[1]][1001:2000,1:1000])

cov_full <- sparsebnUtils::as.sparse(cbind( rbind(cov_expr_methyl[[1]][1:1000,1:1000], x),rbind(t(x), d)))


cov_full_inv = solve(cov_full)

```
