---
title: "gene expr - cnv"
output: html_notebook
---

```{r setup}
require(tidyverse)
require(fst)
require(furrr)
plan(multiprocess)
```

# join gene expr and cnv data

Load the prepared (see unmangle_data.R) data sets for CNV + gene expression data.

```{r paged.print=TRUE}
Lung_GeneExpression <- read_fst("../../data/intermediates/Lung_GeneExpression_normal.fst") %>% as_data_frame
Lung_CNV <- read_fst("../../data/intermediates/Lung_CNV_transposed.fst") %>% as_data_frame

```

The full covariance matrix is too large - select the 1000 variables with highest std deviation for each data set.

```{r}
Lung_CNV[,-1] %>% 
  apply(MARGIN = 2,FUN = sd) -> cnv_vars
Lung_GeneExpression[,-1] %>% 
  apply(MARGIN = 2,FUN = sd) -> expr_vars
```
```{r}
cnv_vars_sel <- names(sort(cnv_vars,decreasing = T)[1:1000])
expr_vars_sel <- names(sort(expr_vars,decreasing = T)[1:1000])
```

Join the resulting subsets, where the cnv data is in .cnv and the gene expr in .geneexpr named variables.

```{r}
genexpr_cnv <- 
  inner_join(
  select(Lung_CNV,SampleID,cnv_vars_sel) %>% rename_at(cnv_vars_sel, .funs = str_c,'.cnv'),
  select(Lung_GeneExpression,SampleID,expr_vars_sel) %>% rename_at(expr_vars_sel, .funs = str_c,'.geneexpr'), by = c("SampleID"))

# genexpr_cnv <- 
  # mutate_at(genexpr_cnv, .vars=vars(starts_with("TCGA")), as.double)

# write_fst(genexpr_cnv, '../../data/intermediates/genexpr_cnv_join.fst' )
```

# correlation networks

First calculate the correlation matrices on the full sample of variables (i.e. cnv / gene expr) for a range of lambda values AND get some idea of reasonable values for lambda.

```{r}
df_data <- genexpr_cnv[,-1]
sparsebnUtils::sparsebnData(df_data, type ="cont") ->dat
nn <- sparsebnUtils::num.samples(dat) # number of samples in the dataset / equivalent to nrow(dat$data)
cnv_columns <- str_which(tbl_vars(df_data), 'cnv$')
geneexpr_columns <-str_which(tbl_vars(df_data),  'geneexpr$')
blacklist <- rbind(expand.grid(cnv_columns,cnv_columns),expand.grid(geneexpr_columns,geneexpr_columns)) %>% as.matrix
lambdas <- sparsebnUtils::generate.lambdas(sqrt(nn), 0.001, lambdas.length = 20, scale = "linear")

cov_exprcnv <- sparsebn::estimate.covariance(dat, blacklist=blacklist, lambdas=lambdas)      
```

Now plot lambda versus the fraction of non-zero coefficients:

```{r}
data_frame(lambda=lambdas[-1], 
              fr_nonzero = map(cov_exprcnv,as.vector) %>% 
                map(`>`,0) %>% 
                map_dbl(mean))-> fr_nonzero_data
  
  
fr_nonzero_data %>% ggplot(aes(lambda,fr_nonzero)) + geom_line() + geom_point()+ theme_minimal()

```

We're still going with lambda = 5.0

```{r}
cov_exprcnv_picked <- cov_exprcnv[[which.min((lambdas-5)**2)]]
```

```{r}
saveRDS(cov_exprcnv_picked, '../../data/intermediates/cov_exprcnv.rds')
write_fst(genexpr_cnv,'../../data/intermediates/genexpr_cnv.fst')
```

# Clustering

```{r}

```



```{r}
genexpr_cnv %>% group_by(SampleID) %>% nest %>% mutate(data=map(data, as.matrix))-> genexpr_cnv_nested
iprd <- expand.grid(1:nrow(genexpr_cnv_nested),1:nrow(genexpr_cnv_nested))
rw <- genexpr_cnv_nested$data

# prd <- cross_df(genexpr_cnv_nested$data, genexpr_cnv_nested$data)
m <- solve(cov_exprcnv_picked)
dstnc <- map2_dbl(iprd[,1],iprd[,2],~tcrossprod( (rw[[..1]]-rw[[..2]]) %*% m, (rw[[..1]]-rw[[..2]])))

```


